Raw Loans Analysis
========================

------

### Load libraries

```{r}
#library(maps)
#library(Hmisc)
```



------

## Processing

Download the data, read the data in and save the raw data as an RDA file

```{r loadData,cache=TRUE}
setwd("C:/Users/Lukasz/Documents/R/AssignmentProject_2/code/rawcode")
getwd()
# download.file("https://spark-public.s3.amazonaws.com/dataanalysis/samsungData.rda",destfile="../../data/samsungDataDownloaded.rda")
# 
# dateDownloaded <- date()
# dateDownloaded
# 
# saveRDS(dateDownloaded,file="../../data/samsungDataDownloaded.rda")
# 
# load("../../data/samsungDataDownloaded.rda")
# 
# save(samsungData, dateDownloaded,file="../../data/samsungData.rda")

load("../../data/samsungData.rda")

```


### Look at the data set

```{r, dependson="loadData"}
head(samsungData)
names(samsungData)

sort(table(names(samsungData)))

unique(samsungData$activity)
unique(samsungData$subject)
table(samsungData$subject)

```

### replace duplicated column names
```{r, dependson="loadData"}
colNames <- names(samsungData)
colNamesTable <- table(colNames)

dupColNames <- names(colNamesTable[colNamesTable > 1])



for (name in colNames) {
  
  if (is.element(name, dupColNames)) {
    name <- paste(name,)
  }
}

```

### Find out about missing values
```{r, dependson="loadData"}
sum(is.na(samsungData))
```

### Create training set and test sets
```{r splitData, dependson="loadData"}

trainData <- samsungData[is.element(samsungData$subject, c(1,3,5,6)),]
table(trainData$subject)

testData <- samsungData[is.element(samsungData$subject, c(27,28,29,30)),]
table(testData$subject)

```

### Try with a tree
```{r , dependson="splitData"}
library(tree)

tree(activity ~ .,data=trainData)

```


### Convert Factors to indexed equivalents.
```{r processData, dependson="cleanData"}

loansProcessed<- loansComplete

loansProcessed$Interest.Rate        <- as.numeric(sub("%","",loansProcessed$Interest.Rate))/100
loansProcessed$Debt.To.Income.Ratio <- as.numeric(sub("%","",loansProcessed$Debt.To.Income.Ratio))/100
# loansFinal$Loan.Length          <- as.numeric(sub(" months","",loansComplete$Loan.Length))
# 
# loansFinal$Employment.Length[loansFinal$Employment.Lenth == "n/a" ] <- NA    
# 
# loansFinal$Employment.Length    <- as.numeric(sub(" years?", "" ,sub("10\\+ years", "10", sub("< 1 year","0",loansFinal$Employment.Length))))

loansProcessed$FICO.Range           <- sapply(strsplit(as.character(loansProcessed$FICO.Range) , "-" ), function(x) mean(as.numeric(x)))

# levels(loansFinal$Loan.Purpose) <- 1:length(levels(loansFinal$Loan.Purpose))
# loansFinal$Loan.Purpose         <- as.numeric(loansFinal$Loan.Purpose)
# 
# 
# levels(loansFinal$State) <- 1:length(levels(loansFinal$State))
# loansFinal$State        <- as.numeric(loansFinal$State)
# 
# levels(loansFinal$Home.Ownership) <- 1:length(levels(loansFinal$Home.Ownership))
# loansFinal$Home.Ownership        <- as.numeric(loansFinal$Home.Ownership)
# 

```

### Filter one outlier
```{r filterData, dependson="processData"}

loansFinal <- loansProcessed[!(loansProcessed$FICO.Range < 700 & loansProcessed$Interest.Rate < 0.065 & loansProcessed$Loan.Length == "60 months"),]

``` 

------

## Exploratory analysis

### Make some univariate plots/summaries

```{r, dependson="processData"}
hist(loansFinal$Interest.Rate, breaks= 100)
lines(density(loansFinal$Interest.Rate[loansFinal$Loan.Length ==36 ]), col ="red")
lines(density(loansFinal$Interest.Rate[loansFinal$Loan.Length ==60 ]), col ="blue")

quantile(loansFinal$Interest.Rate)
hist(loansFinal$FICO.Range,breaks=unique(loansFinal$FICO.Range))
quantile(loansFinal$FICO.Range)
hist(loansFinal$Employment.Length)
hist(loansFinal$Amount.Requested)
hist(loansFinal$Monthly.Income, breaks=100)
hist(loansFinal$Revolving.CREDIT.Balance, breaks=100)
hist(loansFinal$Revolving.CREDIT.Balance, breaks=100)

```

### Make some univariate tables
```{r, dependson="processData"}
table(loansFinal$Employment.Length)
table(loansFinal$Loan.Purpose)
table(loansFinal$State)
table(loansFinal$Employment.Length)
table(loansFinal$Loan.Purpose)
table(loansFinal$Home.Ownership)


```


### IntrestRates vs FICO -> strong dependence
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$FICO.Range, varwidth=T)

```

### IntrestRates vs Loan length -> strong dependence
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Loan.Length, varwidth=T)
```

### IntrestRates vs Ammount Requested
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Amount.Requested, varwidth=T)
```

### IntrestRates vs Employment Dur
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Employment.Length, varwidth=T)
```

### FICO vs Employment Dur
```{r, dependson="ProcessData"}
boxplot(loansFinal$FICO.Range ~ loansFinal$Employment.Length, varwidth=T)
```

### FICO vs Earnings
```{r, dependson="ProcessData"}
boxplot(loansFinal$FICO.Range ~ loansFinal$Monthly.Income, varwidth=T)
```

### FICO vs State
```{r, dependson="ProcessData"}
boxplot(loansFinal$FICO.Range ~ loansFinal$State, varwidth=T)
```

### IntrestRates vs Home Ownership
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Home.Ownership, varwidth=T)
```

### IntrestRates vs open credit lines
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Open.CREDIT.Lines, varwidth=T)
```

### IntrestRates vs revolving Credit Balance
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Revolving.CREDIT.Balance, varwidth=T)
```

### IntrestRates vs inquiries
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Inquiries.in.the.Last.6.Months, varwidth=T)
```

### IntrestRates vs loan  purpose -> medium dependence
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Loan.Purpose, varwidth=T)
```

### IntrestRates vs debt to income
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Debt.To.Income.Ratio, varwidth=T)
```



### try to find correlation for variables when FICO range is set constant 
```{r, dependson="loadData"}


ficoLevels <- sort(unique(loansFinal$FICO.Range))
corrCoeff <- NULL

i <- 1

for(fico in ficoLevels ) {
  
  loansFICO <- loansFinal[loansFinal$FICO.Range == fico,]
  
  loansFICO$FICO.Range <- NULL
  
  corrCoeff <-  cbind(corrCoeff, cor(loansFICO)[,"Interest.Rate"])

  i <- i+1
}    

colnames(corrCoeff) <- ficoLevels
    
```


------

## Modeling


### IntrestRates vs debt to income
```{r, dependson="ProcessData"}
boxplot(loansFinal$Interest.Rate ~ loansFinal$Debt.To.Income.Ratio, varwidth=T)
lmDTIR <- lm(loansFinal$Interest.Rate ~ loansFinal$Debt.To.Income.Ratio)
plot(loansFinal$Interest.Rate ~ loansFinal$Debt.To.Income.Ratio, pch=19, col="blue")
abline(a=lmDTIR$coeff[1], b=lmDTIR$coeff[2], col="red")

summary(lmDTIR)

anova(lmDTIR)
```

### IntrestRates vs FICO -> two distinctive dependencies groups
```{r, dependson="ProcessData"}

lmFICO <- lm(loansFinal$Interest.Rate ~ loansFinal$FICO.Range)
plot(loansFinal$Interest.Rate ~ loansFinal$FICO.Range, pch=19, col= as.numeric(as.factor(loansFinal$Loan.Length)))

abline(a=lmFICO$coeff[1], b=lmFICO$coeff[2], col="red")

summary(lmFICO)

anova(lmFICO)
```

### IntrestRates vs FICO -> two distinctive dependencies groups
```{r, dependson="ProcessData"}

lmLP <- lm(loansFinal$Interest.Rate ~ loansFinal$Loan.Purpose)
boxplot(loansFinal$Interest.Rate ~ loansFinal$Loan.Purpose, pch=19,)

abline(a=lm$coeff[1], b=lmFICO$coeff[2], col="red")

summary(lmLP)

anova(lmLP)
```

### FICO vs debt to income
```{r, dependson="ProcessData"}

lmFICO_DTIR <- lm(loansFinal$FICO.Range ~ loansFinal$Debt.To.Income.Ratio)

plot(loansFinal$FICO.Range~ loansFinal$Debt.To.Income.Ratio, pch=19,)

abline(a=lmFICO_DTIR$coeff[1], b=lmFICO_DTIR$coeff[2], col="red")

summary(lmFICO_DTIR)

anova(lmFICO_DTIR)
```

### FICO vs monthly income
```{r, dependson="ProcessData"}

lmFICO_MI <- lm(loansFinal$FICO.Range ~ loansFinal$Monthly.Income)

plot(loansFinal$FICO.Range ~ loansFinal$Monthly.Income, pch=19,)

abline(a=lmFICO_MI$coeff[1], b=lmFICO_MI$coeff[2], col="red")

summary(lmFICO_MI)

anova(lmFICO_MI)
```

### Interest Rate vs monthly income
```{r, dependson="ProcessData"}

lmMI <- lm(loansFinal$Interest.Rate ~ loansFinal$Monthly.Income)

plot(loansFinal$Interest.Rate ~ loansFinal$Monthly.Income, pch=19,)

abline(a=lmMI$coeff[1], b=lmMI$coeff[2], col="red")

summary(lmMI)

anova(lmMI)
```

### Interest Rate vs monthly income
```{r, dependson="ProcessData"}

lmS <- lm(loansFinal$Interest.Rate ~ loansFinal$State)

plot(loansFinal$Interest.Rate ~ loansFinal$State, pch=19,)


summary(lmS)

anova(lmS)
```



### Interest Rate vs Inq
```{r, dependson="ProcessData"}

lmINQ <- lm(loansFinal$Interest.Rate ~ loansFinal$Inquiries.in.the.Last.6.Months)

boxplot(loansFinal$Interest.Rate ~ loansFinal$Inquiries.in.the.Last.6.Months, varwidth=T)

abline(a=lmINQ$coeff[1], b=lmINQ$coeff[2], col="red")

summary(lmINQ)

anova(lmINQ)
```


### Final Models
```{r finalModels, dependson="filterData"}

lmFinal <- lm(loansFinal$Interest.Rate ~ loansFinal$FICO.Range + 
                                          loansFinal$Loan.Length + 
                                          loansFinal$Inquiries.in.the.Last.6.Months + 
                                          loansFinal$Loan.Length*loansFinal$FICO.Range 
                                          )


summary(lmFinal)

anova(lmFinal)

ind.36 <- loansFinal$Loan.Length == "36 months"
ind.60 <- loansFinal$Loan.Length == "60 months" 


loansFinal.36 <- loansFinal[ind.36,]

loansFinal.60 <- loansFinal[ind.60,]

```

### 3D Plots
```{r finalModels, dependson="filterData"}
library(scatterplot3d)

par(mfrow=c(1,2))

s3d <- scatterplot3d(x = loansFinal.36$FICO.Range, y=  loansFinal.36$Inquiries.in.the.Last.6.Months, z =loansFinal.36$Interest.Rate, pch = 20, main = "36 months loans", ylim = c(0,8), xlab = "FICO", ylab ="Inquiries in the Last 6 Months", zlab = "Interest Rate (%)")

s3d$plane3d(c(lmFinal$coef[1],lmFinal$coeff[2], lmFinal$coeff[4]) , lty = "dotted")

orig <- s3d$xyz.convert(loansFinal.36$FICO.Range, loansFinal.36$Inquiries.in.the.Last.6.Months, loansFinal.36$Interest.Rate)
plane <- s3d$xyz.convert(loansFinal.36$FICO.Range, loansFinal.36$Inquiries.in.the.Last.6.Months, fitted(lmFinal)[ind.36])
i.negpos <- 1 + (resid(lmFinal)[ind.36] > 0)
segments(orig$x, orig$y, plane$x, plane$y,col = c("blue", "red")[i.negpos])



s3d <- scatterplot3d(x = loansFinal.60$FICO.Range, y=  loansFinal.60$Inquiries.in.the.Last.6.Months, z =loansFinal.60$Interest.Rate, pch = 20, main = "60 months loans",xlim=c(600,850), ylim = c(0,8) , xlab = "FICO", ylab ="Inquiries in the Last 6 Months", zlab = "Interest Rate")

s3d$plane3d(c(lmFinal$coef[1]+lmFinal$coef[3],lmFinal$coeff[2]+lmFinal$coef[5], lmFinal$coeff[4]) , lty = "dotted")

orig <- s3d$xyz.convert(loansFinal.60$FICO.Range, loansFinal.60$Inquiries.in.the.Last.6.Months, loansFinal.60$Interest.Rate)
plane <- s3d$xyz.convert(loansFinal.60$FICO.Range, loansFinal.60$Inquiries.in.the.Last.6.Months, fitted(lmFinal)[ind.60])
i.negpos <- 1 + (resid(lmFinal)[ind.60] > 0)
segments(orig$x, orig$y, plane$x, plane$y,col = c("blue", "red")[i.negpos])


```


### Plots
```{r, dependson="FinalModels"}

par(mfrow=c(1,2))

plot(loansFinal$FICO.Range, loansFinal$Interest.Rate, col = as.factor(loansFinal$Loan.Length), pch=19, xlab="FICO score", ylab ="Interest Rate (%)")

legend(700,0.26, c("36 months loan", "60 months loan", "model for 36 months loan", "model for 60 months loan"), pch=19, col = as.factor(unique(loansFinal$Loan.Length)))

abline(a=lmFinalFICO$coefficients[1], b=lmFinalFICO$coefficients[2] , col="orange", lwd=3 )

abline(a=lmFinalFICO$coefficients[1] + lmFinalFICO$coefficients[3], b=lmFinalFICO$coefficients[2] + lmFinalFICO$coefficients[4]  , col="blue", lwd=3 )

abline(a=lmFinal$coefficients[1], b=lmFinalFICO$coefficients[2] , col="orange", lwd=3 )

abline(a=lmFinal$coefficients[1] + lmFinalFICO$coefficients[3], b=lmFinalFICO$coefficients[2] + lmFinalFICO$coefficients[4]  , col="blue", lwd=3 )



plot(loansFinal$Inquiries.in.the.Last.6.Months, loansFinal$Interest.Rate, col = as.factor(loansFinal$Loan.Length), pch=19, ylab="Interest Rate (%)", xlab ="Inquiries in the Last 6 Months")


abline(a=lmFinalINQ$coefficients[1], b=lmFinalINQ$coefficients[2] , col="orange", lwd=3 )
abline(a=lmFinalINQ$coefficients[1] + lmFinalINQ$coefficients[3], b=lmFinalINQ$coefficients[2] , col="blue", lwd=3 )
```


